---
title: "Differential Expression Analysis - Tutorial"
author: Rosa De Santis
date: November 30, 2022
output:
  html_notebook:
    toc: true
    toc_float: true
    collapsed: false
    theme: cerulean
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=F,message=FALSE,warning=FALSE)
```

```{r message=FALSE,warning=FALSE}
#installing packages
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("edgeR")
install.packages("statmod")
install.packages("knitr")
install.packages("reshape2")
install.packages("ggplot2")
install.packages("pheatmap")
install.packages("ghibli")
install.packages("enrichR")
install.packages("gplots")
```

```{r message=FALSE,warning=FALSE}
#Libraries loading
library(edgeR)
library(limma)
library(pheatmap)
library(stringr)
library(gplots)
library(enrichR)
library(ggplot2)
library(ghibli)
library(RColorBrewer)
library(reshape2)
library(statmod)

#Data loading
count_matrix=read.delim("TP63_transdifferentiation_count_matrix.txt", header=T) #load count matrix
sample_info=read.delim("TP63_transdifferentiation_metadata.txt", header=T) #load metadata about samples
```

# Samples description

Analysis of differentially expressed genes between cells infected with GFP (controls) and WT TP63-KLF4 (treatment). The number of replicates in this study are:

```{r}
tab=as.data.frame(table(sample_info$condition)) #summary of replicates
colnames(tab)=c("Condition","Samples")

knitr::kable(tab)
```

# Normalization {.tabset}

## Normalized tables

```{r}
cpm <- cbind(count_matrix[,1:2],cpm(count_matrix[,-c(1,2)])) #CPM normalization
knitr::kable(head(cpm))

lcpm <- cbind(cpm[,1:2],log2(cpm[,-c(1,2)]+1)) #log2 transformation on pseudo-counts to control variance
knitr::kable(head(lcpm))
```

## Effect of normalization

```{r}
count_toplot=count_matrix[,-2]
count_toplot=melt(count_toplot,id.vars="ensembl_gene_id")
count_toplot=count_toplot[count_toplot$value != 0,]
ggplot(count_toplot,aes(x=variable,y=value,fill=variable)) + theme_classic() + geom_boxplot(outlier.shape = NA) + ylim(0,500)

cpm_toplot=cbind(count_matrix[,1:2],cpm(count_matrix[,-c(1,2)])) #CPM normalization
cpm_toplot=cpm_toplot[,-2]
cpm_toplot=melt(cpm_toplot,id.vars="ensembl_gene_id")
cpm_toplot=cpm_toplot[cpm_toplot$value != 0,]
ggplot(cpm_toplot,aes(x=variable,y=value, fill=variable)) + theme_classic() + geom_boxplot(outlier.shape = NA) + ylim(0,100)

```

# Data exploration {.tabset}

## Visualization of know markers

```{r}
genes=c("TP63", "KRT14") #selection of known markers

data=cpm[cpm$external_gene_name%in%genes,-c(1)] #extract genes from table
data=melt(data,id.vars="external_gene_name") #prepare data for plotting
data=merge(data,sample_info,by.x="variable",by.y="sample_name")


ggplot(data, aes(x=external_gene_name, y=value, fill=condition)) + scale_fill_manual(values=c("green2", "darkorchid2"))+
    geom_boxplot(width=0.4,size=.2) + theme_classic()+theme(axis.title = element_blank()) + facet_wrap(.~external_gene_name,scales = "free")

```

## PCA

```{r}
pca=prcomp(t(lcpm[,-c(1,2)])) #apply PCA on log2-transformed counts
percentVar=pca$sdev^2/sum(pca$sdev^2)
d=merge(pca$x, sample_info, by.x=0, by.y="sample_name")
attr(d, "percentVar")=percentVar


  ggplot(d, aes(x=d[,2], y=d[,3], fill=condition)) + geom_point(color="black", alpha=.7,size=5,pch=21) +
    theme_bw()+
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.text = element_text(size=10),
          axis.title = element_text(size=10),
          legend.title = element_text(size=10))+
    labs(fill="Condition")+scale_fill_manual(values=c("green2", "darkorchid2"))+
    xlab(paste0("PC1: ", round(attr(d, "percentVar")[1] *100), "% variance"))+
    ylab(paste0("PC2: ", round(attr(d, "percentVar")[2] *100), "% variance"))
```

## HCL

```{r}
sampleDistMatrix <- cor(lcpm[,-c(1,2)]) #perform Pearson correlation on log2-transformed counts
ann=data.frame(row.names=sample_info$sample_name,condition=sample_info$condition)
  
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
annCol<-list(condition=c(GFP="green2", WT="darkorchid2"))
pheatmap(sampleDistMatrix,
           col=colors, show_colnames = F,show_rownames=T, annotation_row = ann, fontsize=5, annotation_colors = annCol)
```

# Differential Expression Analysis {.tabset}

## Filter out lowly expressed genes

```{r}
counts=data.frame(row.names = count_matrix[,1],count_matrix[,-c(1,2)])
dge<-DGEList(counts=counts, group = factor(sample_info$condition))

keep<-filterByExpr(y=dge) #filter out lowly expressed genes
dge<-dge[keep,,keep.lib.sizes=FALSE]

knitr::kable(data.frame(Filtering=c("Before","After"),Genes=c(nrow(counts),nrow(dge$counts))))

```

## DEA table

```{r}
dge <- calcNormFactors(object=dge) #normalization to perform DEA
dge <- estimateDisp(y = dge, robust = TRUE)

factor=sample_info$condition
design <- model.matrix(~0 + factor)
rownames(design) <- sample_info$sample_name
colnames(design) <- gsub("factor","",colnames(design))
fit <- glmQLFit(dge, design)
dea <- glmQLFTest(fit,contrast = c(-1,1))


res <- topTags(dea, n=nrow(dea$table), adjust.method="fdr")$table #Get DEA table
res$topDE <- ifelse(res$logFC > 1.5 & res$FDR < 0.05, "Up", "NDE") #Mark UP genes
res$topDE <- ifelse(res$logFC < (-1.5) & res$FDR < 0.05, "Down", res$topDE) #Mark DOWN genes
res=merge(res,count_matrix[,1:2],by.x=0,by=1) #Put Symbol names
res=res[order(res$logFC,decreasing = T),] #Order by LFC

de_table=as.data.frame(table(res$topDE))
colnames(de_table)=c("Deregulation","Genes")
knitr::kable(de_table)
head(res)
```

## Volcano Plot

```{r}
ghibli_colors <- ghibli_palette("KikiMedium", type = "discrete") #check palettes list on 

ghibli_subset <- c(ghibli_colors[3], "grey80", ghibli_colors[4]) #Put the colors for Up and Down
res$topDE=factor(res$topDE,levels=c("Up","NDE","Down"))

ggplot(data=res, aes(x=logFC, y=-log10(FDR), color = topDE)) + 
  geom_point() +
  theme_minimal() +
  scale_colour_discrete(type = ghibli_subset, breaks = c("Up", "Down"))+
  geom_hline(yintercept = -log10(0.05),linetype="dashed",color="red4")+
  geom_vline(xintercept = 1.5,linetype="dashed",color="red4")+
  geom_vline(xintercept = -1.5,linetype="dashed",color="red4")+
  xlim(c(-(max(abs(res$logFC))),(max(abs(res$logFC)))))


```

# Pathway Analysis

```{r fig.height=10}
dbs <- c("KEGG_2021_Human","GO_Biological_Process_2021", "ENCODE_and_ChEA_Consensus_TFs_from_ChIP-X")

enrichedup <- enrichr(res[res$topDE=="Up","external_gene_name"], dbs)
enricheddown <- enrichr(res[res$topDE=="Down","external_gene_name"], dbs)


test=list()
for(i in 1:length(dbs)){
 up=enrichedup[[i]]
 up$type="up"
 up=up[order(up$Adjusted.P.value),] 
 down=enricheddown[[i]]
 down$type="down"
 down=down[order(down$Adjusted.P.value),] 
 gos=rbind(down[1:5,],up[1:5,])
 gos$Term=factor(gos$Term, levels=unique(as.character(gos$Term)))
 gos$Database=dbs[i]
 test[[i]]=gos
 
}

merge <- data.table::rbindlist(test)
#merge$Term <- factor(merge$Term, levels=make.unique(as.character(merge$Term), sep = "_"))
merge$count=as.numeric(sapply(strsplit(as.character(merge$Overlap),"/"), "[[", 1))
merge$total=as.numeric(sapply(strsplit(as.character(merge$Overlap),"/"), "[[", 2))
merge$perc=merge$count/merge$total*100
merge=merge[order(merge$Overlap, decreasing = T),]
merge$Term= factor(merge$Term, levels=unique(as.character(merge$Term)))
  
p=ggplot(merge, aes(x=Term, y=type , label=Adjusted.P.value)) + 
  geom_point(aes(size = perc , color = -log10(Adjusted.P.value)))  +
  labs(title= "EnrichR") + coord_flip() + 
  theme_bw(base_family="Times") + theme(strip.text=element_text(size=10), text = element_text(size=15), axis.text = element_text(size=15)) + labs( x = "Databases ", y=" ")

p + facet_grid(rows=vars(Database), scales="free_y")

```
